---
title: "ST558_Project3_EDA"
author: "Stephanie Shuai"
format: html
editor: visual
---

# Introduction

## Data Description

The Water Potability dataset contains water quality metrics for 3,276 water samples. Each observation includes nine physicochemical properties of the water along with a binary label indicating whether the water is safe for human potable or not.

## Purpose

The goal of this exploratory data analysis is to understand the distributions and relationships among the water quality variables, particularly as they relate to Potability(our response variable). We want to identify which physicochemical properties differ between potable and non-potable water and assess whether any strong patterns or correlations exist. This EDA will inform the development of a classification model to predict water potability based on these features.

# DATA

## Load Libraries

```{r}
library(tidyverse)
library(readr)
library(psych)
library(tidymodels)
```

## Import the Data

We use relative path to import the data.

```{r}
water <- read_csv("water_potability.csv", show_col_types = FALSE)
```

Let's take a first look at the data to understand its structure and dimensions.

```{r}
head(water)
```

```{r}
str(water)
```

```{r}
dim(water)
```

The dataset has 3,276 observations and 10 variables. \# Checking the Data

## Step 1: Check for Missingness

Let's examine the number and percentage of missing values for each variable.

```{r}
sum_na <- function(column) {
  sum(is.na(column))
}

na_counts <- water |>
  summarize(across(everything(), sum_na))

na_counts
```

We can see `ph`, `Sulfate`, and `Trihalomethanes` contain missing values.

## Step 2: Check Column Types and Values

Let's verify that all column types. All predictor variables should be numeric, and `Potability` should be treated as a factor.

```{r}
glimpse(water)
```

The summary statistics: - `ph` ranges from 0 to 14. The mean (7.08) and median (7.04) are very close, suggesting a roughly symmetric distribution. - `Solids` and `Conductivity` has a wide range, and the mean exceeds the median, indicating a right skew. - `Hardness`,`Chloramines`, `Sulfate`, `Trihalomethanes`, `Organic_carbon`, and `Turbidity` have narrower range with mean close to their medians, suggesting symmetric distribution.

Let's also check the unique values of `Potability` before converting it to a factor:

```{r}
unique(water$Potability)
```

`Potability` contains only 0 and 1, confirming it is a binary variable.

## Step 3: Rename Variables and Convert Types

Let's rename all variables to lower snake_case for consistency using `rename_with` and convert `Potability` into a factor.

```{r}
water <- water |> 
  rename_with(tolower) |>
  mutate(
    potability = factor(potability, levels = c(0,1), labels = c("Not Potable", "Potable"))
  )
```

Let's verify the changes:

```{r}
str(water)
```

```{r}
table(water$potability)
prop.table(table(water$potability))
```

We can see that the classes are imbalanced, with approximately 61% of samples are not potable and 39% are potable.

# Summarizations

Let's produce meaningful summary statistics and visualizations to explore the relationships in the data.

## Summary Statistics: Overall

Let's compute key descriptive statistics for each numeric predictor. This gives us a baseline understanding of the center, spread, and range of each variables.

```{r}
water |>
  select(where(is.numeric)) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  group_by(variable) |>
  summarise(
    n = sum(!is.na(value)),
    mean = round(mean(value, na.rm = TRUE), 2),
    median = round(median(value, na.rm = TRUE), 2),
    sd = round(sd(value, na.rm = TRUE), 2),
    min = round(min(value, na.rm = TRUE), 2),
    max = round(max(value, na.rm = TRUE), 2)
  )
```

## Summary Statistics: Grouped by Potability

We want to see if the distribution of the predictors differ between potable and non-potable water.

```{r}
water |>
  select(where(is.numeric), potability) |>
  pivot_longer(-potability, names_to = "variable", values_to = "value") |>
  group_by(potability, variable) |>
  summarise(
    n = sum(!is.na(value)),
    mean = round(mean(value, na.rm = TRUE), 2),
    median = round(median(value, na.rm = TRUE), 2),
    sd = round(sd(value, na.rm = TRUE), 2),
    .groups = "drop"
  ) |>
  arrange(variable, potability)
```

## Visualization

### Distribution of the Response Variable

```{r}
ggplot(water, aes(x = potability, fill = potability)) + 
  geom_bar() +
  labs(
    title = "Distribution of Water Potability",
    x = "Potability",
    y = "Count"
  ) +
  scale_fill_manual(values = c("Not Potable" = "#E74C3C", "Potable" = "#2ECC71")) +
  theme_minimal() +
  theme(legend.position = "none")
```

### Histogram of All Numeric Predictors

Let's visualize the distribution of each predictor, colored by potability class.

```{r, warning=FALSE}
water |>
  select(where(is.numeric), potability) |>
  pivot_longer(-potability, names_to = "variable", values_to = "value") |>
  ggplot(aes(x = value, fill = potability)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  labs(
    title = "Distribution of Predictors by Potability",
    x = "Value",
    y = "Count",
    fill = "Potability"
  ) +
  scale_fill_manual(values = c("Not Potable" = "#E74C3C", "Potable" = "#2ECC71")) +
  theme_minimal() 

```

### Boxplots by Potability

Boxplots allow us to compare the central tendency and spread of each predictor across potability groups.

```{r,warning=FALSE}
water |>
  select(where(is.numeric), potability) |>
  pivot_longer(-potability, names_to = "variable", values_to = "value") |>
  ggplot(aes(x = potability, y = value, fill = potability)) +
  geom_boxplot(alpha = 0.7)+
  facet_wrap(~ variable, scales = "free_y", ncol = 3) +
  labs(
    title = "Boxplots of Predictors by Potability",
    x = "Potability",
    y = "Value",
  ) +
  scale_fill_manual(values = c("Not Potable" = "#E74C3C", "Potable" = "#2ECC71")) +
  theme_minimal() 
```

### Correlation Analysis

Let's examine the correlations between all numeric predictors.

```{r}
water_cor <- water |>
  mutate(potability_num = as.numeric(potability) - 1) |>
  select(where(is.numeric)) |>
  cor(use = "pairwise.complete.obs")

round(water_cor, 2)

water_count_cor <- tibble(
  variable = rownames(water_cor),
  correlation_with_potability = water_cor[, "potability_num"]) |>
  filter(variable != "potability_num") |>
  arrange(desc(abs(correlation_with_potability))) 

water_count_cor
```

# Split the data

We will use the `tidymodels` framework to split the data into training and testing sets, and create cross-validation folds on the training sets.

## Handle Missing Values Before Splitting

We impute missing values with the column mean using `tidyr::replace_na()`.

```{r}
water_clean <- water |>
  replace_na(list(
    ph = mean(water$ph, na.rm = TRUE),
    sulfate = mean(water$sulfate, na.rm = TRUE),
    trihalomethanes = mean(water$trihalomethanes, na.rm = TRUE)
  ))
```

```{r}
colSums(is.na(water_clean))
```

## Train/Test Split(75/25, Stratified on Potability)

We use `intial_split` from `tidymodels` with the`strate` argument to ensure that the proportion of potable vs. non-potable is preserved in both the training and testing sets.

```{r}
set.seed(123)

water_split <- initial_split(water_clean, prop = 0.75, strata = potability)
water_train <- training(water_split)
water_test <- testing(water_split)

water_train |>
  count(potability) |>
  mutate(prop = n / sum(n))

water_test |>
  count(potability) |>
  mutate(prop = n / sum(n))

```

Both sets maintain approximately the same proportion of portable vs. non-portable samples,\
confirming that the stratified split worked correctly.

## 10-Fold Cross-Validation

On the training set, we create a 10-fold cross-validation split. This will be used for model tuning and evluation.

```{r}
set.seed(123)

water_folds <- vfold_cv(water_train, v = 10, strata = potability)
water_folds
```
