---
title: "ST558_Project3_EDA"
author: "Stephanie Shuai"
format: html
editor: visual
---

# Introduction

## Data Description

The Water Potability dataset contains water quality metrics for 3,276 water samples.
Each observation includes nine physicochemical properties of the water along with a binary label indicating whether the water is safe for human potable or not.

The variables in this dataset are:

1.  `ph`: pH value of the water (0 to 14).

2.  `Hardness`: Capacity of water to precipitate soap in mg/L.

3.  `Solids`: Total dissolved solids in ppm.

4.  `Chloramines`: Amount of Chloramines in ppm.

5.  `Sulfate`: Amount of Sulfates dissolved in mg/L.

6.  `Conductivity`: Electrical conductivity of water in μS/cm.

7.  `Organic_carbon`: Amount of organic carbon in ppm.

8.  `Trihalomethanes`: Amount of Trihalomethanes in μg/L.

9.  `turbidity`: Measure of light emitting property of water in NTU.

10. `Potability`: Indicates if water is safe for human consumption. (1 = Potable, 0 = Not Potable)

## Purpose

The goal of this exploratory data analysis is to understand the distributions and relationships among the water quality variables, particularly as they relate to Potability(response variable). We want to identify which physicochemical properties differ between potable and non-potable water and assess whether any strong patterns or correlations exist. This EDA will inform the development of a classification model to predict water potability based on these features.

------------------------------------------------------------------------

# Data

## Load Libraries

```{r}
library(tidyverse)
library(readr)
library(tidymodels)
```

## Import the Data

We use relative path to import the data.

```{r}
water <- read_csv("water_potability.csv", show_col_types = FALSE)
```

Let's take a first look at the data to understand its structure and dimensions.

```{r}
head(water)
```

```{r}
str(water)
```

```{r}
dim(water)
```

The dataset has 3,276 observations and 10 variables.

------------------------------------------------------------------------

# Checking the Data

## Check for Missingness

Let's examine the number and percentage of missing values for each variable.

```{r}
sum_na <- function(column) {
  sum(is.na(column))
}

na_counts <- water |>
  summarize(across(everything(), sum_na))

na_counts
```

We can see `ph`, `Sulfate`, and `Trihalomethanes` contain missing values.

## Check Column Types and Values

Let's verify that all column types. All predictor variables should be numeric, and `Potability` should be treated as a factor.

```{r}
glimpse(water)
```

The summary statistics:

-   `ph` ranges from 0 to 14. The mean (7.08) and median (7.04) are very close, suggesting a roughly symmetric distribution.

-   `Solids` and `Conductivity` has a wide range, and the mean exceeds the median, indicating a right skew. -

-   `Hardness`,`Chloramines`, `Sulfate`, `Trihalomethanes`, `Organic_carbon`, and `Turbidity` have narrower range with mean close to their medians, suggesting symmetric distribution.

Let's also check the unique values of `Potability` before converting it to a factor:

```{r}
unique(water$Potability)
```

`Potability` contains only 0 and 1, confirming it is a binary variable.

## Rename Variables and Convert Types

Let's rename all variables to lower snake_case for consistency using `rename_with` and convert `Potability` into a factor.

```{r}
water <- water |> 
  rename_with(tolower) |>
  mutate(
    potability = factor(potability, levels = c(0,1), labels = c("Not Potable", "Potable"))
  )
```

Let's verify the changes:

```{r}
str(water)
```

```{r}
table(water$potability)
prop.table(table(water$potability))
```

We can see that the classes are imbalanced, with approximately 61% of samples are not potable and 39% are potable.

------------------------------------------------------------------------

# Summarizations

Let's produce meaningful summary statistics and visualizations to explore the relationships in the data.

## Summary Statistics: Overall

Let's compute key descriptive statistics for each numeric predictor. This gives us a baseline understanding of the center, spread, and range of each variables.

```{r}
water |>
  select(where(is.numeric)) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  group_by(variable) |>
  summarise(
    n = sum(!is.na(value)),
    mean = round(mean(value, na.rm = TRUE), 2),
    median = round(median(value, na.rm = TRUE), 2),
    sd = round(sd(value, na.rm = TRUE), 2),
    min = round(min(value, na.rm = TRUE), 2),
    max = round(max(value, na.rm = TRUE), 2)
  )
```

A few things stands out from this table:

-   The variables are on very different scales.
-   For most variables, the mean and median are very close, indicating roughly symmetric distributions. `Solids` is a notable exception, which confirming a right skewed distribution with outliers.

## Summary Statistics: Grouped by Potability

We want to see if the distribution of the predictors differ between potable and non-potable water.

```{r}
water |>
  select(where(is.numeric), potability) |>
  pivot_longer(-potability, names_to = "variable", values_to = "value") |>
  group_by(potability, variable) |>
  summarise(
    n = sum(!is.na(value)),
    mean = round(mean(value, na.rm = TRUE), 2),
    median = round(median(value, na.rm = TRUE), 2),
    sd = round(sd(value, na.rm = TRUE), 2),
    .groups = "drop"
  ) |>
  arrange(variable, potability)
```

The means and medians for most variables are nearly identical across the two potability group,. This pattern repeats across all predictors, suggesting that no individual predictor serves as a reliable separator, and our models will need to rely on combinations of features.

------------------------------------------------------------------------

## Visualization

### Distribution of the Response Variable

```{r}
ggplot(water, aes(x = potability, fill = potability)) + 
  geom_bar() +
  labs(
    title = "Distribution of Water Potability",
    x = "Potability",
    y = "Count"
  ) +
  scale_fill_manual(values = c("Not Potable" = "#E74C3C", "Potable" = "#2ECC71")) +
  theme_minimal() +
  theme(legend.position = "none")
```

The dataset is imbalanced. There are 1,998 samples of non-potable compared to 1278 of potable samples.

### Histogram of All Numeric Predictors

Let's visualize the distribution of each predictor, colored by potability class.

```{r, warning=FALSE}
water |>
  select(where(is.numeric), potability) |>
  pivot_longer(-potability, names_to = "variable", values_to = "value") |>
  ggplot(aes(x = value, fill = potability)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  labs(
    title = "Distribution of Predictors by Potability",
    x = "Value",
    y = "Count",
    fill = "Potability"
  ) +
  scale_fill_manual(values = c("Not Potable" = "#E74C3C", "Potable" = "#2ECC71")) +
  theme_minimal() 

```

The histogram shows that potable and non-potable water samples have nearly identical distributions across all nine predictors. Most variables follow a bell-shaped distribution for both classes. Solids is the exception, displaying a right skewed shape with a long tail. 

### Boxplots by Potability

Boxplots allow us to compare the central tendency and spread of each predictor across potability groups.

```{r,warning=FALSE}
water |>
  select(where(is.numeric), potability) |>
  pivot_longer(-potability, names_to = "variable", values_to = "value") |>
  ggplot(aes(x = potability, y = value, fill = potability)) +
  geom_boxplot(alpha = 0.7)+
  facet_wrap(~ variable, scales = "free_y", ncol = 3) +
  labs(
    title = "Boxplots of Predictors by Potability",
    x = "Potability",
    y = "Value",
  ) +
  scale_fill_manual(values = c("Not Potable" = "#E74C3C", "Potable" = "#2ECC71")) +
  theme_minimal() 
```
Across all nine predictors, the median, IQR, and whisker lengths are nearly identically between potable and non-potable. `solids` stands out with a wider spread and more outliers, but both classes behave identically.

### Correlation Analysis

Let's examine the correlations between all numeric predictors and their relationship with potability. 

```{r}
water_cor <- water |>
  mutate(potability_num = as.numeric(potability) - 1) |>
  select(where(is.numeric)) |>
  cor(use = "pairwise.complete.obs")

round(water_cor, 2)
```
The correlations between all nine predictors are very small. This indicates low multicollinearity, meaning each predictors carries largely independent information. 


```{r}
water_count_cor <- tibble(
  variable = rownames(water_cor),
  correlation_with_potability = water_cor[, "potability_num"]) |>
  filter(variable != "potability_num") |>
  arrange(desc(abs(correlation_with_potability))) 
  
water_count_cor
```
None of the predictors have a pratically meaningful linear relationship with water potability. 

------------------------------------------------------------------------

# Key EDA Takeaways

## Based on the summaries and visualizations above:

1.  Missing values was found in `ph`, `sulfate`, and `Trihalomethanes`.

2.  The response variable`potability` is imbalanced.

3.  Visualizations show nearly identical distributions between two classes across all numeric variables.

4.  The correlation analysis shows weak linear relationships between predictors and `potability`.

[Click here for the Modeling Page]()
