---
title: "ST558_Project3_Modeling"
author: "Stephanie Shuai"
format: html
editor: visual
---

# Introduction

In our exploratory data analysis, we investigated the water potability dataset, which contains 3,276 water samples and 10 variables. The goal of this analysis is to build a model of predicting `potability` (whether water is safe for human consumption).

Our EDA revealed that the target variables is imbalanced and that individual predictors do not linearly separate the classes well. Therefore, we will rely on non-linear algorithms using the `tidymodels` framework. We will evaluate two models: a **Classification Tree** and a **Random Forest**, comparing them based on their logarithmic loss(`log-loss`).

# Setup and Data Preparation

```{r}
library(tidyverse)
library(tidymodels)
library(tree)
```

First, we load the raw dataset, convert their column names to lowercase, and format the response variable `potability` as a factor.

```{r}
water <- read_csv("water_potability.csv", show_col_types = FALSE) |>
  rename_with(tolower) |>
  mutate(
    potability = factor(potability, levels = c(0,1), labels = c("Not Potable", "Potable"))
  )
```

# Data Splitting

We will use the `tidymodels` framework to split the data into training and testing sets, and create cross-validation folds on the training sets.

## Handle Missing Values Before Splitting

We impute missing values with the column mean using `tidyr::replace_na()`.

```{r}
water_clean <- water |>
  replace_na(list(
    ph = mean(water$ph, na.rm = TRUE),
    sulfate = mean(water$sulfate, na.rm = TRUE),
    trihalomethanes = mean(water$trihalomethanes, na.rm = TRUE)
  ))
```

```{r}
colSums(is.na(water_clean))
```

## Train/Test Split(70/30, Stratified on Potability)

We use `initial_split` from `tidymodels` with the`strata` argument to ensure that the proportion of potable vs. non-potable is preserved in both the training and testing sets.

```{r}
set.seed(123)

# 70/30 split
water_split <- initial_split(water_clean, prop = 0.70, strata = potability)
water_train <- training(water_split)
water_test <- testing(water_split)

# Check proportions
water_train |>
  count(potability) |>
  mutate(prop = n / sum(n))

water_test |>
  count(potability) |>
  mutate(prop = n / sum(n))

```

Both sets maintain approximately the same proportion of potable vs. non-potable samples, confirming that the stratified split worked correctly.

## 5-Fold Cross-Validation

On the training set, we create a 5-fold cross-validation split. This will be used for model tuning and evaluation.

```{r}
set.seed(123)

water_folds <- vfold_cv(water_train, v = 5, strata = potability)
water_folds
```

## Recipe

Our EDA revealed missing values in `ph`, `sulfate`, and `trihalomethan`. Instead of dropping these rows, we will impute the missing values with the mean of each column.

```{r}
water_rec <- recipe(potability ~., data = water_train) |>
  step_impute_mean(ph, sulfate, trihalomethanes) |>
  step_normalize(all_numeric_predictors())
```

------------------------------------------------------------------------

# Model 1: Classification Tree

## Understanding Classification Trees

A **Classification Tree** is a supervised learning algorithm that works by repeatedly splitting the data into distinct groups based on the predictor variables.

# Random Forest
