---
title: "ST558_Project3_Modeling"
author: "Stephanie Shuai"
format: html
editor: visual
---

# Introduction

In our exploratory data analysis, we investigated the water potability dataset, which contains 3,276 water samples and 10 variables. The goal of this analysis is to build a model of predicting `potability` (whether water is safe for human consumption).

Our EDA revealed that the target variables is imbalanced and that individual predictors do not linearly separate the classes well. Therefore, we will rely on non-linear algorithms using the `tidymodels` framework. We will evaluate two models: a **Classification Tree** and a **Random Forest**, comparing them based on their logarithmic loss(`log-loss`).

# Setup and Data Preparation

```{r}
library(tidyverse)
library(tidymodels)
library(rpart)
library(ranger)
```

First, we load the raw dataset, convert their column names to lowercase, and format the response variable `potability` as a factor.

```{r}
water <- read_csv("water_potability.csv", show_col_types = FALSE) |>
  rename_with(tolower) |>
  mutate(
    potability = factor(potability, levels = c(0,1), labels = c("Not Potable", "Potable"))
  )
```

# Data Splitting

We will use the `tidymodels` framework to split the data into training and testing sets, and create cross-validation folds on the training sets.

## Handle Missing Values Before Splitting

We impute missing values with the column mean using `tidyr::replace_na()`.

```{r}
water_clean <- water |>
  replace_na(list(
    ph = mean(water$ph, na.rm = TRUE),
    sulfate = mean(water$sulfate, na.rm = TRUE),
    trihalomethanes = mean(water$trihalomethanes, na.rm = TRUE)
  ))
```

```{r}
colSums(is.na(water_clean))
```

## Train/Test Split(70/30, Stratified on Potability)

We use `initial_split` from `tidymodels` with the`strata` argument to ensure that the proportion of potable vs. non-potable is preserved in both the training and testing sets.

```{r}
set.seed(123)

# 70/30 split
water_split <- initial_split(water_clean, prop = 0.70, strata = potability)
water_train <- training(water_split)
water_test <- testing(water_split)

# Check proportions
water_train |>
  count(potability) |>
  mutate(prop = n / sum(n))

water_test |>
  count(potability) |>
  mutate(prop = n / sum(n))
```

Both sets maintain approximately the same proportion of potable vs. non-potable samples, confirming that the stratified split worked correctly.

## 5-Fold Cross-Validation

On the training set, we create a 5-fold cross-validation split. This will be used for model tuning and evaluation.

```{r}
set.seed(123)

water_folds <- vfold_cv(water_train, v = 5, strata = potability)
water_folds
```

## Recipe

Our EDA revealed missing values in `ph`, `sulfate`, and `trihalomethan`. Instead of dropping these rows, we will impute the missing values with the mean of each column.

```{r}
water_rec <- recipe(potability ~., data = water_train) |>
  step_impute_mean(ph, sulfate, trihalomethanes) |>
  step_normalize(all_numeric_predictors())
```

------------------------------------------------------------------------

# Model 1: Classification Tree Model

## Understanding Classification Tree Model

A **Classification Tree** is a supervised learning algorithm that works by repeatedly splitting the data into distinct groups based on the predictor variables.

## Tuning the Classification Tree Model

We will fit a classification treee using the `rpart` engine, leaving `cost_complexity` open for tuning.

```{r}
# Specify the model
tree_spec <- decision_tree(
  cost_complexity = tune(),
  tree_depth = tune()
) |>
  set_engine("rpart") |>
  set_mode("classification")

# Create the workflow
tree_wkf <- workflow() |>
  add_recipe(water_rec) |>
  add_model(tree_spec)   

# Create a tuning grid
tree_grid <- grid_regular(
  cost_complexity(),
  tree_depth(),
  levels = 5
)
```

Now we tune the model across our 5-fold cross-validation sets. We will evulate performance using **log-loss** (which measures the accuracy of predicted probabilities).

```{r}
# Define our metric set (Log-Loss)
eval_metrics <- metric_set(mn_log_loss)

# Tune the model
set.seed(123)
tree_res <- tune_grid(
  tree_wkf,
  resamples = water_folds,
  grid = tree_grid,
  metrics = eval_metrics
)
```

Let's look at the top 5 hyperparameter combinations based on lowest log-loss:

```{r}
tree_res |>
  collect_metrics() |>
  filter(.metric == "mn_log_loss") |>
  arrange(mean)
```

# Random Forest Model

A **Random Forest** is an ensemble method that builds upon the classification tree. Instead of training one tree, it trains hundreds of trees. To ensure the trees are diverse, each tree is trained on a random sample of the data, and at each split, it only considers a random subset of the predictors.

We will tune three parameters:

-   `mtry`: The number of predictors randomly sampled at each split.

-   `trees`: The total number of trees to grow.

-   `min_n`: The minimum number of data points in a node required to split it further.

```{r}
# Specify the model
rf_spec <- rand_forest(
  mtry = tune(),
  trees = tune(),
  min_n = tune()
) |>
  set_engine("ranger") |>
  set_mode("classification")

# Create the workflow
rf_wkf <- workflow() |>
  add_recipe(water_rec) |>
  add_model(rf_spec)

# Create a tuning grid 
rf_grid <- grid_regular(
  mtry(range = c(2,7)),
  trees(range = c(100, 300)),
  min_n(range = c(5, 20)),
  levels = 3
)
```

Tune the random forest:
```{r}
set.seed(123)
rf_res <- tune_grid(
  rf_wkf,
  resamples = water_folds,
  grid = rf_grid,
  metrics = eval_metrics
)

rf_res |>
  collect_metrics() |>
  filter(.metric == "mn_log_loss") |>
  arrange(mean)
```

-----------

# Final Model Selection

We will now compare the best log-loss scores from both models to select the final winner. 
```{r}
best_tree_logloss <- tree_res |>
  show_best(metric = "mn_log_loss", n = 1) |>
  pull(mean)

best_rf_logloss <- rf_res |>
  show_best(metric = "mn_log_loss", n = 1) |>
  pull(mean)

round(best_tree_logloss,3)
round(best_rf_logloss,3)
```

The results show that the **Random Forest**(log-loss = 0.613) outperforms the Classification Tree(log-loss=0.666). Since lower log-loss indicates better probability calibration, the Random Forest is the winner. 

## Final Fit and Evaluation

We will finalize the random forest workflow with its optimal hyperparameters, fit it to the entire training set, and evaluate it one last time on the untouched test set.

```{r}
# Select best parameters
best_rf_params <- rf_res |>
  select_best(metric = "mn_log_loss")

# Finalize workflow
final_rf_wkf <- rf_wkf |>
  finalize_workflow(best_rf_params)

# Perform final fit
final_fit <- final_rf_wkf |>
  last_fit(water_split, metrics = metric_set(accuracy, mn_log_loss))

# View test set metrics
final_fit |>
  collect_metrics()
```

## Save Model for API Deployment

We extract the fully trained workflow (fitted on the complete training set) and save it as an `.rds` file. This file will be loaded by our `plumber` API script to serve predictions inside our Docker container. 

```{r}
# Extract the finalized, fitted workflow
best_model <- final_fit |>
  extract_workflow()

# Save the model 
saveRDS(best_model, "best_model.rds")
```

